Refractory anaphylaxis: Data from the European Anaphylaxis Registry

### 2018-08-15 analysis/paper/paper.Rmd
berfore changing after a while

---
title: "Refractory anaphylaxis: Data from the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik*
  - Sabine Dölle-Bierke*
  - Franziska Ruëff**
  - Margitta Worm*
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

@@

**Keywords:** 
anaphylaxis, adrenaline (epinaphrine), betablockers, Insect venom allergy, Drug allergic reactions

# Abstract: 
<!-- 350 Words unstructured -->
Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 mg adrenaline) is a rare and often fatal hypersensitivity reaction. Based on this, definition, its prevalence, and risk factors are only vaguely described.  
Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we were able to identify cases of refractory anaphylaxis (n = `r v$casesR`) and to analyse these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).  Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r elicitExact[4,3]`% vs `r elicitExact[4,2]`%, p = `r elicitExact[4,4]`). The refractory cases occurred more frequently in a perioperative setting (`r v$perioperative[3]`% vs. `r v$perioperative[2]`, p < 0.0001). Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases was given intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001). Second line treatment options (e.g. vasopression with dopamine, methylene blue, glucagon) were not used at all in the treatment of refractory cases. The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).
Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these standards can be improved. The low use of intramuscular adrenaline as a first line therapy might contribute to refractoriness of an episode. Also, the lack of suggested second line medication use in refractory cases indicate that the availability of these drug is restricted or anaphylaxis management guidelines may need a revision in order to provide clear algorithms for severe refractory anaphylaxis.

<!--
# Introduction: 
The mainstay of anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might insufficient to restore a stable patient status.

Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which is unresponsive to the treatment with at least 2 doses of minimum 300 mg adrenaline without the expected normalization of clinical symptoms.

# Methods
We aimed to distinguish the prevalence of refractory anaphylaxis among anaphylaxis cases and factors which may increase the risk of a refractory anaphylaxis episode. 
Using the Data from the European Anaphylaxis Registry we selected cases where patients received at least two doses of adrenalin, and failed to recover adequately. The flowchart in figure \@ref(fig:flowchart) represents the detailed cases selection process.

The final database consisted of `r sum(countries)` cases from `r length(countries)` countries: `r paste0(names(countries),": ",countries,collapse=", ")`.

@@

## Demographics
The European Anaphylaxis Registry contained `r {data3[,"rANA"]=="yes"} %>% which %>% length` cases of refractory anaphylaxis and in `r {data3[,"rANA"]=="no"} %>% which %>% length` cases severe, non-refractory anaphlaxis was diagnosed. The mean age at the reaction was `r {data3$d_age[data3[,"rANA"]=="yes"]} %>% mean(na.rm=T) %>% signif(3)` ± `r data3$d_age[data3[,"rANA"]=="yes"] %>% sd(na.rm=T)` years, and did not differ from severe, non-refractory cases, p = `r t.test(data3$d_age[data3[,"rANA"]=="yes"],data3$d_age[data3[,"rANA"]=="no"])$p.value %>% signif(3)`). There were however significantly more cases of concomitant malignant diseases in the refractory anaphylaxis group. The demographic summary of refractory cases can be found in table \@ref(tab:demoTab).

## Elicitors
Refractory anaphylaxis was most commonly elicited by drugs (significantly more often than in severe, non-refractory cases), followed by food and insects (table \@ref(tab:elicitorTab)). 
The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics and X-ray contrast medium. Insect-venom refractory cases were significantly more often elicited by bee-stings than severe, non-refractory cases, although the yellow-jacket-stings were most common in other anaphylaxis cases (\@ref(tab:elicitExact)).

## Therapy
Adrenaline  iv. as a first-line treatment of anaphylaxis was significantly more often given in refractory cases (`r therapyTab[2,3]`% vs `r therapyTab[2,3]`%).

@@

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

@@

```{r demoTab}
knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. ")
```

```{r elicitorTab}
knitr::kable(elicitorTab, caption = "Summary of elicitors in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]"))
```

```{r elicitExact}
knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of elicitors in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"))
```

* p-value derived from the Fischer exact test.

```{r therapyTab}
knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"))
```

* p-value derived from the Fischer exact test.

>>>>

---
title: "Refractory anaphylaxis: Data from the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik*
  - Sabine Dölle-Bierke*
  - Franziska Ruëff**
  - Margitta Worm*
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: expert-review-of-clinical-immunology.csl # Insert path for the bib-style
#abstract: 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

```{r}
periopAna <- data3$rANA[which(data3$q_152_location=="medical practice, hospital")] %>% 
  as.factor %>%  summary()

```

@@

**Keywords:** 
anaphylaxis, adrenaline (epinephrine), beta-blockers, Insect venom allergy, Drug allergic reactions

# Abstract
<!-- 350 Words unstructured -->
Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 mg adrenaline) is a rare and often fatal hypersensitivity reaction. Based on this, its definition, prevalence, and risk factors are only vaguely described.  
Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we were able to identify cases of refractory anaphylaxis (n = `r v$casesR`) and to analyse these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).  Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r elicitExact[4,3]`% vs `r elicitExact[4,2]`%, p = `r elicitExact[4,4]`). The refractory cases occurred more frequently in a perioperative setting E(`r v$perioperative[3]`% vs. `r v$perioperative[2]`, p < 0.0001). Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases was given intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001). Second line treatment options (e.g. vasopression with dopamine, methylene blue, glucagon) were not used at all in the treatment of refractory cases. The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).
Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these standards can be improved. The low use of intramuscular adrenaline as a first line therapy might contribute to refractoriness of an episode. Also, the lack of suggested second line medication use in refractory cases indicate that the availability of these drug is restricted or anaphylaxis management guidelines may need a revision in order to provide clear algorithms for severe refractory anaphylaxis.

# Introduction

Anaphylaxis is a non-homogeneous clinical diagnosis, depending on various triggering, and augmenting factors. This variability introduces a wide range of possible reaction-symptom severity. Therefore multiple sub-types of anaphylaxis have been previously identified (i.e. food dependent exercise induced anaphylaxis, venom anaphylaxis, biphasic anaphylaxis).

The mainstay of anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might be insufficient to restore a stable patient status. Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 mg adrenaline, does not result in an expected normalization of clinical symptoms. Typical elicitors and symptoms of refractory anaphylaxis, as well as the therapeutic strategy for the most severe cases differ from the usual reactions and call for a specific research and targeted guideline development for refractory anaphylaxis cases.

We aimed to distinguish the prevalence of refractory anaphylaxis among anaphylaxis cases and to describe its symptoms and factors which may increase the risk of a refractory anaphylaxis episode.

# Methods
The European Anaphylaxis Registry (described in detail elsewhere [@Grabenhenrich2016]) provided data for this analysis status from May 2018.  Based on  an electronic questionnaire (available at www.anaphylaxie.net), we selected cases where patients received at least two doses of adrenaline, and failed to recover adequately and assigned them to the "refractory anaphylaxis group". The flowchart in figure \@ref(fig:flowchart) represents the detailed cases selection process.

The final database consisted of `r sum(countries)` cases of refractory anaphylaxis from `r length(countries)` countries: `r paste0(names(countries),": ",countries,collapse=", ")`. We compared these to a group of severe, non-refractory cases of anaphylaxis. Severe reactions were identified based on the definition of by NIAID/FAAN [@Sampson2006] and presented with significant hypoxia, cyanosis, hypotension (SBP < 90 mm Hg in adults), confusion, collapse, loss of conciusness, or incontinence. 
We compared the frequency of various elicitors, symptoms, and factors known to incease the risk of severe anphylaxis [@Worm2018] in both groups. We also compared how the cases were managed.

<!-- ## Data handling and statisitcs -->
The statistical analysis was performed in the R Statistical Package [@R]. Simple comparison of nominal variables was performed using Fischer's exact test, continous variables were analysed using Mann-Whithney U test. We defined statistical significance as α = 0.05. Data along with the analysis script can be accessed here...

@@

<!-- ## Demographics -->

The European Anaphylaxis Registry contained `r {data3[,"rANA"]=="yes"} %>% which %>% length` cases of refractory anaphylaxis (reported in years 2007-2017) and in `r {data3[,"rANA"]=="no"} %>% which %>% length` cases severe, non-refractory anaphylaxis was diagnosed. The frequency of refractory anaphylaxis was therefore `r ((data3$rANA=="yes") %>% which() %>% length / (data3$d_year_reaction %in% 2007:2017) %>% which() %>% length ) %>% {.*100} %>% signif(2)`% of all anaphylaxis cases reported in the registry from European countries in years 2007 - 2017. It may be stated that each year `r table(data3$d_year_reaction, data3$rANA) %>% {cbind(.[,1]/(.[,1]+.[,2]),.[,2]/(.[,1]+.[,2]))[which(rownames(.) %in% 2007:2017),2]} %>% {paste0(mean(.) %>% signif(3)*100, "% $\\pm$ ", sd(.) %>% signif(3)*100,"%")}` of severe anaphylactic episodes are refractory to treatment with adrenaline. Moreover, if we consider patients undergoing a medical procedure (who undergo anaphylaxis in a perioperative setting), nearly `r (periopAna[2]/(periopAna[2]+periopAna[1])) %>% signif(3)*100`% (`r (periopAna[2]/(periopAna[2]+periopAna[1]) %>% signif(3)*100 / 0.853) %>% signif(3)` times more) patients present with reactions that do not respond to adrenaline.

The percentage of males within the  refractory anaphylaxis group was `r (as.numeric(demoTab[2,2])/sum(as.numeric(demoTab[2,2])+as.numeric(demoTab[1,2]))*100) %>% {round(.,1)}`% and the mean age at the reaction was `r {data3$d_age[data3[,"rANA"]=="yes"]} %>% mean(na.rm=T) %>% signif(3)` ± `r data3$d_age[data3[,"rANA"]=="yes"] %>% sd(na.rm=T) %>% signif(3)` years, which did not differ from severe, non-refractory cases, p = `r t.test(data3$d_age[data3[,"rANA"]=="yes"],data3$d_age[data3[,"rANA"]=="no"])$p.value %>% signif(3)`).  There were however significantly more cases of concomitant malignant diseases in the refractory anaphylaxis group. Patients with refractory reactions more frequently had a previous anaphylactic reaction in their medical history (p = `r data3$q_410_prev_v5 %>% table(data3$rANA) %>%{.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% signif(3)`).
The demographic summary of refractory cases can be found in table \@ref(tab:demoTab)

<!-- # ## Elicitors -->

```{r}
locTab <- data3$q_152_location %>% table(data3$rANA)
 # data3$rANA[which(data3$q_152_location=="medical practice, hospital")] %>% as.factor %>%  summary()
```

Refractory anaphylaxis was most commonly elicited by drugs (significantly more often than in severe, non-refractory cases), followed by food and insects (table \@ref(tab:elicitorTab)). In `r elicitorTab[1,8]`% of food elicited cases the diagnosis of food allergy was previously confirmed. Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
Patients with refractory anaphylaxis experienced the reaction while undergoing a medical procedure in `r (locTab[3,2]/sum(locTab[,2])*100) %>% {signif(.,3)}`% of the times (vs. `r (locTab[3,1]/sum(locTab[,1])*100) %>% {signif(.,3)}` in severe, non-refractory cases, p < 0.0001)

The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics and X-ray contrast medium. Insect-venom-elicited, refractory cases were significantly more often elicited by bee-stings than severe, non-refractory cases, whereas the yellow-jacket-stings were most common in non-refractory anaphylaxis cases (table \@ref(tab:elicitExact)).

<!-- ## Symptoms -->
Milder anaphylaxis symptoms (i.e. pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) showed significantly more often in severe anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were associated more often with the refractory anaphylaxis cases. Table \@ref(tab:symptTab) summarizes the most prominent differences in anaphylaxis symptoms among both groups.

<!-- ## Therapy -->
When evaluating the therapeutic procedures, adrenaline iv. as a first-line treatment of anaphylaxis was significantly more often given in refractory cases (`r therapyTab[2,3]`% vs. `r therapyTab[1,3]`%, p < 0.0001) it has been also given more often as a second line therapy in refractory cases.

```{r}
med.time2 <-data3$q_530_adren2_time_in_min_v5 %>% as.character() %>% as.numeric %>% split(data3$rANA) %>% lapply(median, na.rm=T)
```

Median time to second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes). <!--  wilcox.test(data3$q_530_adren2_time_in_min_v5[data3$rANA=="yes"],data3$q_530_adren2_time_in_min_v5[data3$rANA=="no"])$p.value %>% signif(3)-->

```{r}
adr2prof <- rcalc("q_530_adren2_who_v5","professional")
```

In `r adr2prof[3]`% of refractory cases, the second dose of adrenaline was given by a professional (significantly more than in non-refractory cases `r adr2prof[2]`%, p < 0.0001).

Corticosteroids iv. were the second most frequently administered group of drugs in refractory cases (as a first and second line treatment), outpacing antihistaminic drugs and volume replacement therapy, and were significantly more often given in refractory cases. Volume replacement therapy was given initially in `r therapyTab[4,3]`% of refractory cases but was sustained only in `r therapyTab[5,3]`% as the therapy progressed in the hospital environment. 
<!--(possible counfounding variable by patients who died!!! they did not receive second line therapy and might need to be discarded from the analysis)-->

Second line medication like dopamine, glucagon and methylene blue were neglected in all refractory anaphylaxis cases as well as in severe non-refractory ones, however patients with refractory anaphylaxis were more frequently admitted to the hospital and treated at the ICUs. Table \@ref(tab:therapyTab) illustrated the therapy of refractory anaphylaxis cases.

<!-- ## Death  -->
Fatal reactions occurred after 30 minutes to 120 minutes - when the CPR was unsuccessful. In cases where patients responded to life support procedures, but failed to be reanimated due to post-resuscitative complication (e.g. hypoxic brain injury), death occured in the next 3-8 days.

<!-- ## Cofactors -->
Patients with refractory anaphylaxis more often had concomitant asthma and malignant diseases in the past. Also other unspecified concomitant conditions were significantly more often reported in refractory cases (although they were not explicitly stated in the questionnaire). It is worth noticing that concomitant cardiologic conditions, diabetes and mastocytosis were similarly frequent in both groups.

Patients with refractory anaphylaxis more often used proton pump inhibitors and acetylsalicilic acid than patients with severe non-refractory anaphylaxis. Other (non-disclosed) medications were also more frequent in refractory cases.

The intensity of of physical exercise exceeding the reaction was indifferent between groups, however the psychological burden was reported 3 times more frequently in refractory cases (see table \@ref(tab:cofactorsTab)).

```{r}
#data3$q_410_other_codisease_free[which(data3$rANA=="yes")]
#rcalc("q_4212_exercise_intensity","moderate eg. vigorous housework or gardening")
#data3$q_423_other_drug_free[which(data3$rANA=="yes")]

```

@@

## Key results

### Frequency and patients at risk:
The frequency of refractory anaphylaxis was less than 1% of severe anaphylaxis cases per year. This suggests that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapeutic strategy, commonly outlined in anaphylaxis management guidelines. Refractory anaphylaxis was most often iatrogenic: ie. elicited by drugs and occurring in a medical practice or a hospital. Perioperative cases of refractoriness have been previously published to be the most common in case series analysis [source]. Therefore patients who undergo medical procedures and have a history of anaphylaxis should remain under special caution.

### Elicitors
Although drugs were the most commonly occurring elicitor, there were also relatively more refractory cases elicited by bee stings. Plausible explanations for this observation, might include either longer exposition to the allergen (longer stings predispose to deeper deposition of the venom and slower metabolizing) or more efficient vasodilation due to biochemical differences between wasp and bee venom. [insekt sting anaphylaxis, Golden, 2007]. In such case - patients might benefit from using vasopressive medication.

Antibiotics, although commonly triggering type IV hypersensitivity reactions, are rarely reported in the literature as the cause of a refractory anaphylaxis. Radio-contrast media (RCM) are however common in eliciting hypersensitivity episodes refractory to adrenaline. We saw multiple refractory reactions to antibiotics but in clinical practice, these might be overestimated, as patients in a perioperative setting receive multiple drugs along with antibiotics.

### Cofactors 
Although we have previously identified older age and male sex to be associated with more severe anaphylaxis, these factors seem to be less important when refractory anaphylaxis is considered. Asthma and malignant diseases in the medical history were more frequent in the refractory anaphylaxis group and the reason for this is unclear. It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients.

After performing a thorough analysis of co-factors associated with an increased risk of severe anaphylaxis [@Worm2018] we saw that concomitant asthma rather had a protective effect. This association is probably non-linear as the present analysis indicated concomitant asthma as a risk factor for a refractory episode. The reason for this is unclear. Either patients receive already antihistaminic and mast cell blocking agents and their reaction to adrenaline is therefore restricted, or the episode elicits a bronchospasm that is refractory to adrenaline and therefore results in a prolonged anaphylactic episode. This would also clarify why we saw more respiratory distress symptoms in the refractory anaphylaxis cases.

We have previously proven the role of acetysalicylic acid in increasing the risk of severe anaphylaxis. The role of proton pump inhibitors (PPI) however has not been identified as a co-factor, and there are only reports on PPI as elicitors of anaphylaxis.

### adrenaline use inconsistent and too low
After revising the therapeutic procedures we saw significantly more adrenaline use in the refractory cases, but it could be used even more extensively in these life-threatening reactions. This observation might indicate that in fact the emergency teams are using these life-saving measures too cautiously due to possible side effects. Methylene blue and vasopressors have been described to successfully restore refractory hypotension, but their actual use in anaphylaxis patients is scarce, although they are mentioned in recognized guidelines on the management of anaphylaxis.

## Limitations and strength
Low number of refractory cases prevented us from analyzing the data with more advanced statistical models, however, our analysis it is the first published report on over 30 refractory anaphylaxis cases.

Definition  of refractory anaphylaxis is not universal, and confirming the refractory anaphylaxis based on the answers to our online questionnaire poses additional difficulties.
If fatal reaction occurs before the second dose of adrenaline could be administered to the patient - it cannot be diagnosed as refractory (as we now require at least two doses of minimum 300 µg adrenaline). although it as well could be refractory.

Comparison of refractory anaphylaxis to patients with severe anaphylaxis gave us the possibility to describe which patients have a higher risk of developing a refractory episode, however we can not answer the question which patients have the higher risk of developing anaphylaxis per se.

## Conclusion

Refractory anaphylaxis is a rare form of life-threatening hypersensitivity reaction with high mortality. We propose to increase the use of second line medication such as methylene blue, vasopressin and glucagon
in cases where 2 doses of adrenaline did not result in rapid normalization of anaphylaxis symptoms.

# Acknowledgements
We thank the members of The European Anaphylaxis Registry.
<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->

@@

```{r demoTab}
knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other as fractions [%]")
```

```{r elicitorTab}
knitr::kable(elicitorTab, caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. ANA - anaphylaxis, p-value derived from the Fisher exact test.",
             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]"))
```

```{r elicitExact}
knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of the elicitors broken down to specific elicitors fro meach group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ANA - anaphylaxis, p-value derived from the Fisher exact test. ",
             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"))
```

p-value derived from the Fisher exact test.

```{r symptTab}
knitr::kable(symptTab[,1:4], caption = "Summary of the symptoms in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. ANA - anaphylaxis, p-value derived from the Fisher exact test. ",
             col.names = c("Symptom","severe ANA [%]","refractory ANA [%]","p value"))
```

```{r therapyTab}
knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control. 2nd line therapy was performed after the initial rescue individual called for a professional medical emergeny team. ANA - anaphylaxis, p-value derived from the Fisher exact test. ",
             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"))
```

```{r cofactorsTab}
knitr::kable(cofactorsTabF[,1:4], caption = "Factors potentially increasing the risk of a severe anaphylaxis investigated in refractory cases. ANA - anaphylaxis, p-value derived from the Fisher exact test. ",
             col.names = c("Factor","severe ANA [%]","refractory ANA [%]","p value"))
```
